Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-28 notes/notes-2/notes-2.tex
Extend notes for talk 2
Add sections about full/faithful functors and subcategories and many
basic examples, mostly about monoids.

@@

Next, we consider a further example of a $\Set$-valued functor, which unlike the previous ones is not forgetful.
\begin{example}
  \label{ex:Mset}
  Let $\cat M$ be a category with a single object $M$, i.e.\ a monoid. Then, a functor $F\from\cat M\to\Set$ is a set $S=F(M)$ together with functions $F(m)\from S\to S$ for $m\from M\to M$ in $\cat M$. Hence, we can introduce the short-hand notation
  \begin{align*}
    m\cdot s=F(m)(s).
  \end{align*}
  The functorality axioms for $F$ state that
  \begin{align*}
    1_M\cdot s&=F(1_M)(s)=1_S(s)=s\\
    (m\of m')\cdot s&=F(m\of m')(s)=(F(m)\of F(m'))(s)=F(m)(F(m')(s))=m\cdot (m'\cdot s)
  \end{align*}
  for all $m,m'\from M\to M$ and $s\in S$.

@@

>>>>

We want to define structure-preserving maps between categories.
Since the morphisms are are most interesting part of a category, such a map should not just operate on objects, but on morphisms to.
In order to preserve the structure, it has to map identity morphisms to identity morphisms and compositions of morphisms to compositions.
This is made precise in the following definition.

@@

The image of such a functor can be thought of as a \textit{shadow} of category $\cat{A}$ in $\cat{B}$.
Sometimes we call the first function \demph{action on paths} and the second function \demph{action on morphisms}.

We can easily define composition of functors and an identity functor on every category.
With those we define a category $\Cat$ that contains categories as objects and functors as morphisms.

Let us look at some examples of functors.
We recommend to check the axioms yourself in order to verify that they indeed satisfy our definition.

\begin{example}
  We define a \demph{powerset functor} $\Set \xrightarrow{\mathcal{P}} \Set$ by
  \begin{align*}
    \mathcal{P}(A) &\coloneqq \{ X \mid X \subseteq A \} \\
    \mathcal{P}(f) &\coloneqq (\{x, y, \ldots\} \mapsto \{f(x), f(y), \ldots\})
  \end{align*}
  for all objects $A$ and $B$ and morphisms $A \xrightarrow{f} B$.
  In a similar way we can define a functor that maps a set $A$ to the set of lists or to the set of multisets with elements in $A$ or to many other types of collections.
  In all those cases the action on morphisms allows us to apply a function element-wise on a collection.
  If you have experience in functional programing, this may seem familiar to you.
\end{example}

In the next few examples we will talk about a vary simple type of algebraic structures with a binary operation that we now define.

\begin{definition}
  A set $M$ is a \demph{monoid} if it has a binary operation ${}*{} \from M \times M \to M$ that satisfies the following axioms.
  \begin{description}
    \item [Associativity] $(x * y) * z = x * (y * z)$ for all $x, y, z \in M$.
    \item [Neutral element] There is an element $e \in M$ such that $e * x = x = x * e$ for all $x \in M$.
  \end{description}
\end{definition}

This is the same as the definition of \emph{groups} except that we don't require inverses.
Similar to homomorphisms on groups, we define homomorphisms on monoids.

\begin{definition}
  Given two monoids $M$ and $N$, a function $M \xrightarrow{f} N$ is a \demph{monoid homomorphism} if it satisfies the axioms
  \begin{itemize}
    \item $f(e) = e$
    \item $f(x * y) = f(x) * f(y)$
  \end{itemize}
  where we write $e$ for the neutral element in both monoids and $*$ for the binary operation in both monoids.
\end{definition}

You might already spot some similarity to the functoriality axioms.
We use this in our next example.

\begin{example}
  There is a one-to-one correspondence between monoids and categories with one object.
  Elements of the monoid correspond to morphisms in the category and the binary operation is interpreted as composition of morphisms.
  With this interpretation, the functoriality axioms are the same as the axioms for monoid homomorphisms, which means that the homomorphisms between two groups are exactly the functors between the representing categories.
\end{example}

\begin{example}
  Monoids together with monoid homomorphisms form a category.
  We can define a functor from the category of monoids to $\Set$, that maps every monoid to its element set and every homomorphism to the undelying function.
  This functor maps into a category that is similar to its domain but with less structure on the objects.
  Hence one can say that the functor \emph{forgets} the monoid structure.
  This can be observed on many functors.
  We call such functors \demph{forgetful}.
  Forgetfulness is an informal concept and depends on our interpretation of the category, but usually it is obvious if you should call a functor forgetful.
\end{example}

\begin{example}
  There is a functor from the category of commutative monoids to the category of general monoids which maps every object and every morphism to itself.
  It is injective on objects and on arrows, which means that we can always say where a resulting object \textit{came from}.
  But we still call it \emph{forgetful} because it maps into a category where in general the objects have less structure.
\end{example}

We saw two examples of functors that forget structure, but there are also functors that do the opposite and create structure.
We will take a look at one of them.

\begin{example}
  Given a set $A$, we define the \demph{free monoid} $A^*$ over $A$ as the set of lists with elements in $A$.
  As binary operation we use the concatenation of lists and as neutral element the empty list.
  If we have a function $A \xrightarrow{f} B$ then we can lift it to a monoid homomorphism $A^* \xrightarrow{f^*} B^*$ that applies $f$ pointwise on a list, i.e.\ $f^*([x, y, \ldots]) = [f(x), f(y), \ldots]$.
  This gives us a functor that maps a set to its free monoid and hence \emph{adds structure}.
  In general we call such a functor a \demph{free functor}.
\end{example}

Next, we consider a further example of a $\Set$-valued functor which, unlike the previous one, is not forgetful.
\begin{example}
  \label{ex:Mset}
  Let $\cat M$ be a category with a single object $M$, i.e.\ a monoid. Then, a functor $F\from\cat M\to\Set$ is a set $S=F(M)$ together with functions $F(m)\from S\to S$ for $m\from M\to M$ in $\cat M$. Hence, we can introduce the short-hand notation
  \begin{align*}
    m\cdot s=F(m)(s).
  \end{align*}
  The functorality axioms for $F$ state that
  \begin{align*}
    1_M\cdot s&=F(1_M)(s)=1_S(s)=s\\
    (m\of m')\cdot s&=F(m\of m')(s)=(F(m)\of F(m'))(s)=F(m)(F(m')(s))=m\cdot (m'\cdot s)
  \end{align*}
  for all $m,m'\from M\to M$ and $s\in S$.

@@

\section{Full and Faithful functors}

The following definition introduces a concept that is similar to injectivity of functors but restricts only the action on morphisms, not on objects.

\begin{definition}
  A functor $\cat{A} \xrightarrow{F} \cat{B}$ is \demph{faithful} if for all objects $A, B \in \cat{A}$ the function
  \begin{align*}
    \cat{A}(A, B) &\to \cat{B}(F(A), F(B)) \\
    f &\mapsto F(f)
  \end{align*}
  is \emph{injective}.
\end{definition}

Note that this definition only speaks about what heapens to morphisms between the \emph{same pair} of objects.
For morphisms $A \xrightarrow{f, g} B$, we can conclude that $F(A) = F(B)$ implies $A = B$ but for a morphism $A \xrightarrow{f} B$ and a morphism $C \xrightarrow{g} D$ between another pair of objects, the faithfulness of $F$ doesn't tell us anything.
Let us consider a functor like in the following graphic as an example.
\[ \begin{tikzpicture}
    \draw (0, 0) rectangle (2, 3);
    \node (A) at (0.5, 2.5) {$A$};
    \node (C) at (1, 0.5) {$C$};
    \node (B) at (1.5, 2.5) {$B$};
    \draw[->] (A) to node[auto, swap] {$f$} (C);
    \draw[->] (B) to node[auto] {$g$} (C);

    \draw (3, 0) rectangle (5, 3);
    \node (X) at (4, 2.5) {$X$};
    \node (Y) at (4, 0.5) {$Y$};
    \draw[->] (X) to node[auto] {$u$} (Y);

    \draw[->, dotted, out=45, in=120] (A) to (X);
    \draw[->, dotted, out=45, in=120] (B) to (X);
    \draw[->, dotted, out=-45, in=-120] (C) to (Y);
  \end{tikzpicture} \]
Assume that a functor $F$ is given by the dotted arrows and maps the morphism $f$ and $g$ both to $u$.
The functor is faithful nevertheless because the morphisms $f$ and $g$ start from different objects.

Analogously to the definition of faithfulness, we define fullness of a functor, which is similar to the concept of surjectivity.

\begin{definition}
  A functor $\cat{A} \xrightarrow{F} \cat{B}$ is \demph{full} if for all objects $A, B \in \cat{A}$ the function
  \begin{align*}
    \cat{A}(A, B) &\to \cat{B}(F(A), F(B)) \\
    f &\mapsto F(f)
  \end{align*}
  is \emph{surjective}.
\end{definition}

\section{Subcategories}

With the definition of full and faithful functors we adapted two set theoretic concepts for our work with categories.
The same can be done with the subset relation.
This time there is one obvious way to do it which is the following definition.

\begin{definition}
  Given two categories $\cat{A}$ and $\cat{B}$, we say that $\cat{A}$ is a \demph{subcategory} of $\cat{B}$ if $\ob(\cat{A}) \subseteq \ob(\cat{B})$ and for all objects $A, B \in \cat{A}$, we have the inclusion $\cat{A}(A, B) \subseteq \cat{B}(A, B)$.
\end{definition}

If we consider the inclusion functor $\cat{A} \to \cat{B}$ that maps every object and every morphism to itself, we can ask the question, whether it is full or faithful.
The faithfulness is obvious, but only certain subcategories have full inclusion functors.
This motivates the definition of full subcategories.

\begin{definition}
  A subcategory $\cat{A}$ of a category $\cat{B}$ is \demph{full} if for all objects $A, B \in \cat{A}$, there is an equality $\cat{A}(A, B) = \cat{B}(A, B)$, not just an inclusion.
\end{definition}

With our definition, a subcategory is full if and only if the inclusion functor is full.
An important property is that a full subcategory is completely determined by the selection of objects that it contains.

As the image of a monoid homomorphism is a monoid and the image of a group homomorphism is a group, one could easily assume that the image of a functor is a category, but in general this is wrong as we can see in this example:
\[ \begin{tikzpicture}
  \draw (0, 0) rectangle (3, 5);
  \node (A) at (1, 4.5) {$A$};
  \node (B) at (1, 2.5) {$B$};
  \node (C) at (2, 2.5) {$C$};
  \node (D) at (2, 0.5) {$D$};
  \draw[->] (A) to node[auto] {$f$} (B);
  \draw[->] (C) to node[auto] {$g$} (D);

  \draw (4, 0) rectangle (7, 5);
  \node (X) at (5, 4.5) {$X$};
  \node (Y) at (5, 2.5) {$Y$};
  \node (Z) at (5, 0.5) {$Z$};
  \draw[->] (X) to node[auto] {$u$} (Y);
  \draw[->] (Y) to node[auto] {$v$} (Z);
  \draw[->, bend left=45] (X) to node[auto] {$v \of u$} (Z);

  \draw[->, dotted, in=120, out=45] (A) to (X);
  \draw[->, dotted, in=120, out=45] (B) to (Y);
  \draw[->, dotted, in=-120, out=-45] (C) to (Y);
  \draw[->, dotted, in=-120, out=-45] (D) to (Z);
\end{tikzpicture} \]
Assume that a functor $F$ is given by the dotted arrows and maps the morphism $f$ to $u$ and $g$ to $v$.
Since the image of $F$ contains $X \xrightarrow{u} Y$ and $Y \xrightarrow{i} Z$, it should also contain $v \of u$ to be a category.
But $F$ does not map any morphism to $v \of u$, thus the image is not a category.
